Practical Optional Types for Clojure

### 2015-02-20 formal-model.tex
java interop

It is common in Clojure to use plain immutable hash maps to represent ad-hoc
objects with a known set of keyword entries. We use \HMapliteral{}
types to track known present, known absent and unknown keyword entries
(figure~\ref{figure:hmapsyntax}).

@@

\subsection{Java interoperability}

\begin{figure*}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \reflectiveexpsyntax{}
            \alt  \localhintsyntax{} \alt \lethintsyntax{}
    &\mbox{Expressions} \\
    \v{} &::=& \ldots \alt {\classvalue{\classhint{}} {\overrightarrow {\classfieldpair{\fld{}} {\v{}}}}}
    &\mbox{Values} \\
    \tatypesyntax{}\\
    \typehintenvsyntax{}\\
    \classtableallsyntax{}
  \end{altgrammar}
  $$
  \caption{Java Interoperability Terms and Types}
  \label{figure:reflectivesyntax}
\end{figure*}

As Clojure supports Java interoperability, Typed Clojure has some
integration with Java's type system. Calls to Java methods from
Clojure can be type checked by Typed Clojure.

We do not support Java Generics for several reasons. Firstly
Clojure programmers already use \emph{type hints} to resolve
Java methods which does not interact with generics. Secondly
it would complicate the type system to need to consider advanced
Generics features like existentials and f-bounded polymorphism
which has complicated other type systems like Scala's.
Finally Java interop is rare in Clojure code, so the extra
work needed would not be worth the effort compared to the number
of new bugs core.typed catches.

Thanks to type erasure, only unparameterised classes, primitive
types and array classes live at runtime. This simplifies 
inference for \emph{tools.analyzer}, which uses type hints
to resolve ambiguities in Java interop like overloaded methods.

\subsection{Clojure type hints}

\begin{figure*}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}
    %&&\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \
    %\ \ \ \ \ \ \ \  \ \ \ \ \ \ \ \ \ \ \ \ \ \ \

    &\mbox{Expressions}
  \end{altgrammar}
  $$
  \caption{Internal Language Non-Reflective Java Interoperability Syntax}
  \label{figure:nonreflectivesyntax}
\end{figure*}

@@

core.typed uses tools.analyzer to obtain a rich abstract syntax tree that includes resolved
reflection, however to tell the full story we include a rewrite relation 
in figure~\ref{figure:rewrite} to eliminate reflection when sufficient type hints are provided.
Figure~\ref{figure:hintinfer} models how type hints are propagated.

@@

\subsection{Handling null}
\label{sec:null}

Probably the most common pain point in Java programming is dealing with \emph{null}.
It is crucial that Typed Clojure deals with \emph{null} intelligently so that
all potentially erroneous interactions with \emph{null} in typed code are caught at compile time.
We also want to represent \emph{null} in a way that is natural to Clojure programmers
and models their reasoning of Clojure code.

Firstly, we separate the concepts of \emph{null} and reference types at the type level.
This is unlike Java, where a reference type, like \emph{java.lang.Number}, includes \emph{null}.
We provide a singleton type \Nil{}, which contains just the value \lstinline|nil|,
Clojure's equivalent of \emph{null}.

Armed with general unions and the type \lstinline|nil|, 
we make Typed Clojure's reference types \emph{non-nullable}.
A Java type like \emph{java.lang.Number} is written \lstinline|(U nil java.lang.Number)|
in Typed Clojure.

Typed Racket already provides what is needed to write such types; indeed this is a repurposing
of a common Typed Racket idiom of approximating an Option or Maybe type by
making a union that includes \emph{\#f}, Racket's only false value.
However, like handling covariant arrays, the subtlety lies in our interactions with foreign
Java code.

A set of conversion rules define how to convert a Java type signature into a null-safe
Clojure version. These rules are designed to be conservative and can be explicitly overridden
by programmers. Method parameters are non-nullable, while return types are nullable.
Field types are nullable. Constructor parameters are non-nullable, and return types
are also non-nullable: it is a Java invariant that all constructors return an instance
of the class they are constructing.

>>>>

It is common in Clojure to use plain immutable hash maps to represent ad-hoc
objects with a known set of keyword entries. We use \HMapliteral{}
types (figure~\ref{figure:hmapsyntax}) to track known present, known absent and unknown keyword entries.

@@

\subsection{Java Interoperability and Type Hints}

\begin{figure*}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \reflectiveexpsyntax{}
            \alt  \localhintsyntax{} \alt \lethintsyntax{}
    &\mbox{Expressions}
  \end{altgrammar}
  $$
  \caption{Java Interoperability Terms and Types}
  \label{figure:reflectivesyntax}
\end{figure*}

Clojure supports interoperability with Java, including the ability to
call constructors, methods and access fields. We model
these operations in a restricted setting without class inheritance,
overloading or Java Generics.

We extend the external language with Java field lookups and calls to
methods and constructors (figure~\ref{figure:nonreflectivesyntax}).
There are no typing or evaluation rules for these syntaxes because
Typed Clojure must be convinced that Java interactions are \emph{non-reflective},
ie., are unambiguously resolved at compile-time. Programmers use
type hinted expressions (figure~\ref{figure:nonreflectivesyntax}) 
to help the compiler eliminate reflection.

\begin{figure*}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}
    &\mbox{Expressions}\\

    \v{} &::=& \ldots \alt {\classvalue{\classhint{}} {\overrightarrow {\classfieldpair{\fld{}} {\v{}}}}}
    &\mbox{Values} \\

    \tatypesyntax{}\\
    \typehintenvsyntax{}\\
    \classtableallsyntax{}
  \end{altgrammar}
  $$
  \caption{Internal Language Reflection Resolution Extensions}
  \label{figure:nonreflectivesyntax}
\end{figure*}

@@

core.typed uses tools.analyzer to obtain a rich abstract syntax tree that includes resolved
reflection, however to tell the full story we include a rewrite relation 
in figure~\ref{figure:rewrite} to eliminate reflection when sufficient type hints are provided.
Figure~\ref{figure:hintinfer} shows the interesting cases in type hint inference.

@@
